{\large Gravitational wave sources in our Galactic backyard:}\\{\normalsize Predictions for BHBH, BHNS and NSNS binaries in LISA

### 2021-07-30 paper/tex/sections/abstract.tex
add hyphen to CE

We present predictions for the properties of the population of Galactic double black holes (BHBHs), black hole neutron stars (BHNSs) and double neutron stars (NSNSs) that will be detectable by the planned space-based gravitational wave detector LISA. We use rapid population synthesis to produce an extensive sample of double compact objects (DCOs) and combine this with an empirically-informed model to distribute them in a Milky Way-like galaxy based on their birth metallicity. We investigate the dependence of our results upon underlying physics assumptions by comparing the results of \nModels{} physics variations that vary assumptions relating to mass transfer, common envelope, supernova and wind mass loss physics. We predict that for a 4(10)-year mission, LISA will (typically) detect about \BHBHFourYear{}(\BHBHTenYear{}) BHBHs, \BHNSFourYear{}(\BHNSTenYear{}) BHNSs, \NSNSFourYear{}(\NSNSTenYear{}) NSNSs. The predicted BHBH rate remains notably consistent under different physics assumptions, whilst in contrast the BHNS and NSNS rates each vary over 3 orders of magnitude. We discuss observable characteristics that could be used to distinguish the aforementioned DCOs from the more numerous double white dwarf population as well as for disentangling the BHBH, BHNS and NSNS populations from each other. We additionally assess the possibility of multi-messenger observations of pulsar populations by combining the capabilities of LISA and SKA.

>>>>

We present predictions for the properties of the population of Galactic double black holes (BHBHs), black hole neutron stars (BHNSs) and double neutron stars (NSNSs) that will be detectable by the planned space-based gravitational wave detector LISA. We use rapid population synthesis to produce an extensive sample of double compact objects (DCOs) and combine this with an empirically-informed model to distribute them in a Milky Way-like galaxy based on their birth metallicity. We investigate the dependence of our results upon underlying physics assumptions by comparing the results of \nModels{} physics variations that vary assumptions relating to mass transfer, common-envelope, supernova and wind mass loss physics. We predict that for a 4(10)-year mission, LISA will (typically) detect about \BHBHFourYear{}(\BHBHTenYear{}) BHBHs, \BHNSFourYear{}(\BHNSTenYear{}) BHNSs, \NSNSFourYear{}(\NSNSTenYear{}) NSNSs. The predicted BHBH rate remains notably consistent under different physics assumptions, whilst in contrast the BHNS and NSNS rates each vary over 3 orders of magnitude. We discuss observable characteristics that could be used to distinguish the aforementioned DCOs from the more numerous double white dwarf population as well as for disentangling the BHBH, BHNS and NSNS populations from each other. We additionally assess the possibility of multi-messenger observations of pulsar populations by combining the capabilities of LISA and SKA.

### 2021-07-30 paper/tex/sections/appendix.tex
add hyphen to CE

\textit{Common Envelope:} A common envelope phase follows dynamically unstable mass transfer and we parameterise this using the $\alpha$-$\lambda$ prescription from \citet{Webbink+1984} and \citet{deKool+1990}. We assume $\alpha = 1$, such that all of the gravitational binding energy is available for the ejection of the envelope. For $\lambda$ we use the fitting formulae from \citet{Xu+2010, Xu+2010a}. We assume that any Hertzsprung gap donor stars that initiate a common envelope phase will not survive this phase due to a lack of a steep density gradient between the core and envelope \citep{Taam+2000, Ivanova+2004, Klencki+2021}. This follows the `pessimistic' common envelope scenario \citep[c.f.][]{Belczynski+2007}. We remove any binaries where the secondary immediately fills its Roche lobe upon the conclusion of the common envelope phase as we treat these as failed common envelope ejections, likely leading to a stellar merger.

@@

\subsection{Model variations} \label{sec:variation_assumptions}
In addition to our fiducial model for the formation of DCOs, we explore \nMinusOneModels{} other models in which we change various aspects of the mass transfer, common envelope, supernova and wind mass loss physics assumptions in order to assess the effect of their uncertainties on the overall double compact object detection rates and distributions. Each of the models varies a single physics assumption (fiducial assumptions are outlined in Section~\ref{app:fiducial_physics}) and these models are outlined in Table~\ref{tab:physics_variations}.

@@

Models \modRangeCE{} focus on altering the common envelope physics. We change the common envelope efficiency parameter to $\alpha_{\rm CE} = 0.1, 0.5, 2.0, 10.0$ in models \modAlphaLowest{}, \modAlphaLow{}, \modAlphaHigh{} and \modAlphaHighest{} respectively. In model \modOpt{}, we relax our restriction that Hertzsprung gap donor stars cannot survive common envelope events, thereby following the `optimistic' common envelope scenario. We combine this with model \modCaseBB{} in model \modCaseBBOpt{}.

@@

\begin{table}[htb]
    \centering
    \begin{tabular}{cl}
        \hline \hline
        Model & Physics Variation \\
        \hline \hline
        \modFid & Fiducial (see Section~\ref{app:fiducial_physics}) \\
        \hline
        \modBetaLow & Fixed mass transfer efficiency of $\beta=0.25$ \\ 
        \modBetaMed & Fixed mass transfer efficiency of $\beta=0.5$  \\ 
        \modBetaHigh & Fixed mass transfer efficiency of $\beta=0.75$ \\ 
        \modCaseBB & Case BB mass transfer is always unstable \\
        \modCaseBBOpt & Model \modCaseBB{} + Model \modOpt{} \\
        \hline
        \modAlphaLowest & CE efficiency parameter $\alpha = 0.1$ \\
        \modAlphaLow & CE efficiency parameter $\alpha = 0.5$ \\
        \modAlphaHigh & CE efficiency parameter $\alpha = 2$   \\
        \modAlphaHighest & CE efficiency parameter $\alpha = 10$   \\
        \modOpt & HG donor stars initiating a CE survive CE \\
        \hline
        \modRapid & Fryer rapid SN remnant mass prescription \\
        \modNSLow & Maximum NS mass is fixed to $2\unit{M_{\rm \odot}}$ \\
        \modNSHigh & Maximum NS mass is fixed to $3\unit{M_{\rm \odot}}$ \\
        \modNoPISN & PISN and pulsational-PISN not implemented \\
        \modSigLow & $\sigma_{\rm{rms}}^{\rm{1D}}=100 \unit{km}{s^{-1}}$ for core-collapse supernova \\  
        \modSigLower & $\sigma_{\rm{rms}}^{\rm{1D}}=30  \unit{km}{s^{-1}}$ for core-collapse supernova \\ 
        \modNoBH & Black holes receive no natal kick \\
        \hline
        \modWRLow & Wolf-Rayet wind factor $f_{\rm WR} = 0.1$ \\
        \modWRHigh & Wolf-Rayet wind factor $f_{\rm WR} = 5.0$ \\
        \hline \hline
    \end{tabular}%
    \caption{A description of the \nModels{} binary population synthesis models used in this study. \modFid{} is the fiducial model, \modRangeMT{} change mass transfer physics, \modRangeCE{} change common envelope physics , \modRangeSN{} change supernova physics and \modRangeML{} change wind mass loss \citep[c.f.][Table 2]{Broekgaarden+2021}.}
    \label{tab:physics_variations}
\end{table}

@@

\begin{figure}[h]
    \centering
    \includegraphics[height=0.85\textheight]{fig14_formation_channels.pdf}
    \caption{Fraction of each DCO type that is formed through different formation channels for all physics variations. Channels are described in detail in \citet{Broekgaarden+2021}. The classic, single core CEE and double core CEE channels all require at least one common envelope event whilst only `only stable' consists of only stable mass transfer and `other' contains the remaining binaries which are mainly formed from `lucky' supernova kicks that shrink the binary. \href{https://github.com/TomWagg/detecting-DCOs-in-LISA/blob/main/paper/figures/fig14_formation_channels.png}{\faFileImage} \href{https://github.com/TomWagg/detecting-DCOs-in-LISA/blob/main/paper/figure_notebooks/formation_channels.ipynb}{\faBook}. \todo{Update R and T}}
    \label{fig:formation_channels}
\end{figure}

>>>>

\textit{Common Envelope:} A common-envelope phase follows dynamically unstable mass transfer and we parameterise this using the $\alpha$-$\lambda$ prescription from \citet{Webbink+1984} and \citet{deKool+1990}. We assume $\alpha = 1$, such that all of the gravitational binding energy is available for the ejection of the envelope. For $\lambda$ we use the fitting formulae from \citet{Xu+2010, Xu+2010a}. We assume that any Hertzsprung gap donor stars that initiate a common-envelope phase will not survive this phase due to a lack of a steep density gradient between the core and envelope \citep{Taam+2000, Ivanova+2004, Klencki+2021}. This follows the `pessimistic' common-envelope scenario \citep[c.f.][]{Belczynski+2007}. We remove any binaries where the secondary immediately fills its Roche lobe upon the conclusion of the common-envelope phase as we treat these as failed common-envelope ejections, likely leading to a stellar merger.

@@

\subsection{Model variations} \label{sec:variation_assumptions}
In addition to our fiducial model for the formation of DCOs, we explore \nMinusOneModels{} other models in which we change various aspects of the mass transfer, common-envelope, supernova and wind mass loss physics assumptions in order to assess the effect of their uncertainties on the overall double compact object detection rates and distributions. Each of the models varies a single physics assumption (fiducial assumptions are outlined in Section~\ref{app:fiducial_physics}) and these models are outlined in Table~\ref{tab:physics_variations}.

@@

Models \modRangeCE{} focus on altering the common-envelope physics. We change the common-envelope efficiency parameter to $\alpha_{\rm CE} = 0.1, 0.5, 2.0, 10.0$ in models \modAlphaLowest{}, \modAlphaLow{}, \modAlphaHigh{} and \modAlphaHighest{} respectively. In model \modOpt{}, we relax our restriction that Hertzsprung gap donor stars cannot survive common-envelope events, thereby following the `optimistic' common-envelope scenario. We combine this with model \modCaseBB{} in model \modCaseBBOpt{}.

@@

\begin{table}[htb]
    \centering
    \begin{tabular}{cl}
        \hline \hline
        Model & Physics Variation \\
        \hline \hline
        \modFid & Fiducial (see Section~\ref{app:fiducial_physics}) \\
        \hline
        \modBetaLow & Fixed mass transfer efficiency of $\beta=0.25$ \\ 
        \modBetaMed & Fixed mass transfer efficiency of $\beta=0.5$  \\ 
        \modBetaHigh & Fixed mass transfer efficiency of $\beta=0.75$ \\ 
        \modCaseBB & Case BB mass transfer is always unstable \\
        \modCaseBBOpt & Model \modCaseBB{} + Model \modOpt{} \\
        \hline
        \modAlphaLowest & CE efficiency parameter $\alpha = 0.1$ \\
        \modAlphaLow & CE efficiency parameter $\alpha = 0.5$ \\
        \modAlphaHigh & CE efficiency parameter $\alpha = 2$   \\
        \modAlphaHighest & CE efficiency parameter $\alpha = 10$   \\
        \modOpt & HG donor stars initiating a CE survive CE \\
        \hline
        \modRapid & Fryer rapid SN remnant mass prescription \\
        \modNSLow & Maximum NS mass is fixed to $2\unit{M_{\rm \odot}}$ \\
        \modNSHigh & Maximum NS mass is fixed to $3\unit{M_{\rm \odot}}$ \\
        \modNoPISN & PISN and pulsational-PISN not implemented \\
        \modSigLow & $\sigma_{\rm{rms}}^{\rm{1D}}=100 \unit{km}{s^{-1}}$ for core-collapse supernova \\  
        \modSigLower & $\sigma_{\rm{rms}}^{\rm{1D}}=30  \unit{km}{s^{-1}}$ for core-collapse supernova \\ 
        \modNoBH & Black holes receive no natal kick \\
        \hline
        \modWRLow & Wolf-Rayet wind factor $f_{\rm WR} = 0.1$ \\
        \modWRHigh & Wolf-Rayet wind factor $f_{\rm WR} = 5.0$ \\
        \hline \hline
    \end{tabular}%
    \caption{A description of the \nModels{} binary population synthesis models used in this study. \modFid{} is the fiducial model, \modRangeMT{} change mass transfer physics, \modRangeCE{} change common-envelope physics , \modRangeSN{} change supernova physics and \modRangeML{} change wind mass loss \citep[c.f.][Table 2]{Broekgaarden+2021}.}
    \label{tab:physics_variations}
\end{table}

@@

\begin{figure}[h]
    \centering
    \includegraphics[height=0.85\textheight]{fig14_formation_channels.pdf}
    \caption{Fraction of each DCO type that is formed through different formation channels for all physics variations. Channels are described in detail in \citet{Broekgaarden+2021}. The classic, single core CEE and double core CEE channels all require at least one common-envelope event whilst only `only stable' consists of only stable mass transfer and `other' contains the remaining binaries which are mainly formed from `lucky' supernova kicks that shrink the binary. \href{https://github.com/TomWagg/detecting-DCOs-in-LISA/blob/main/paper/figures/fig14_formation_channels.png}{\faFileImage} \href{https://github.com/TomWagg/detecting-DCOs-in-LISA/blob/main/paper/figure_notebooks/formation_channels.ipynb}{\faBook}. \todo{Update R and T}}
    \label{fig:formation_channels}
\end{figure}

### 2021-07-30 paper/tex/sections/introduction.tex
add hyphen to CE

For the purposes of this investigation, we consider the `classical' isolated binary evolution channel \citep[e.g.][]{Tutukov+1973,Tutukov+1993,Smarr+1976,Srinivasan+1989,Kalogera+2007,Belczynski+2016} in which double compact objects are formed following common envelope ejection or a phase of highly non-conservative mass transfer \citep{Heuvel+2011, vandenHeuvel+2017}. We do not, however, account for several alternative proposed formation channels, which could affect the rate and distribution of detectable NS and BH binaries in LISA. These channels include: dynamical formation in dense star clusters \citep[e.g.][]{Sigurdsson+1993,PortegiesZwart+2000,Miller+2009,Rodriguez+2015}, young/open star clusters \citep[e.g.][]{Ziosi+2014, DiCarlo+2020, Rastello+2020, Rastello+2021} and (active) galactic nuclei discs \citep[e.g.][]{Morris+1993, Antonini+2016, McKernan+2020}, isolated (hierarchical) triple evolution involving Kozai-Lidov oscillations \citep[e.g.][]{Stephan+2016, Silsbee+2017,Antonini+2017, Toonen+2020},  and chemically homogenous evolution through efficient rotational mixing \citep[e.g.][]{deMink+2009,Mandel+2016,Marchant+2016,Marchant+2017,duBuisson+2020}.

>>>>

For the purposes of this investigation, we consider the `classical' isolated binary evolution channel \citep[e.g.][]{Tutukov+1973,Tutukov+1993,Smarr+1976,Srinivasan+1989,Kalogera+2007,Belczynski+2016} in which double compact objects are formed following common-envelope ejection or a phase of highly non-conservative mass transfer \citep{Heuvel+2011, vandenHeuvel+2017}. We do not, however, account for several alternative proposed formation channels, which could affect the rate and distribution of detectable NS and BH binaries in LISA. These channels include: dynamical formation in dense star clusters \citep[e.g.][]{Sigurdsson+1993,PortegiesZwart+2000,Miller+2009,Rodriguez+2015}, young/open star clusters \citep[e.g.][]{Ziosi+2014, DiCarlo+2020, Rastello+2020, Rastello+2021} and (active) galactic nuclei discs \citep[e.g.][]{Morris+1993, Antonini+2016, McKernan+2020}, isolated (hierarchical) triple evolution involving Kozai-Lidov oscillations \citep[e.g.][]{Stephan+2016, Silsbee+2017,Antonini+2017, Toonen+2020},  and chemically homogenous evolution through efficient rotational mixing \citep[e.g.][]{deMink+2009,Mandel+2016,Marchant+2016,Marchant+2017,duBuisson+2020}.

### 2021-07-30 paper/tex/sections/results.tex
add hyphen to CE

We find that detectable BHNSs have even more unequal mass ratios, such that the distribution is almost entire disjoint from that of NSNSs. Moreover, the mass ratio distribution is bimodal, where the two peaks arise from two distinct formation scenarios. Around two thirds of detectable BHNSs experience at least one common envelope event, whilst the last third are formed through only stable mass transfer. The first peak at $q = 0.18$ is from systems that experience at least one common envelope and occurs at the expected mass ratio, which approximately follows the mean BH mass ($\sim 6.5 \unit{M_\odot}$) and NS mass ($\sim 1.2 \unit{M_\odot}$). Yet we also see a second peak at higher mass ratios around $q = 0.34$, which arise from the fraction of the population that underwent only stable mass transfer phases. The stability of mass transfer is strongly dependent on the mass ratio and thus these systems have a bias for more equal mass ratio systems, leading to a peak at higher $q$.

@@

The peak at recent times is due to the fact that most binaries in our sample are formed with merger times on the order of a couple of Gyr (since most form through a common envelope phase that significantly tightens the binary). For a binary to be in the LISA band and detectable it must be near the end of its inspiral and so LISA will be most sensitive to sources that were formed a couple of Gyr ago.

@@

In the lower panel of Fig.~\ref{fig:detectable_distance_dist}, we show the density distribution for detectable DCOs in the galaxy. We see that most sources are concentrated in the Galactic center, with a bias towards to the position of the Sun. Yet we also show that systems are detectable out to the edges of the galaxy. This is especially true for NSNSs, as we highlight by plotting the 75\% contour for the individual DCO types.

>>>>

We find that detectable BHNSs have even more unequal mass ratios, such that the distribution is almost entire disjoint from that of NSNSs. Moreover, the mass ratio distribution is bimodal, where the two peaks arise from two distinct formation scenarios. Around two thirds of detectable BHNSs experience at least one common-envelope event, whilst the last third are formed through only stable mass transfer. The first peak at $q = 0.18$ is from systems that experience at least one common-envelope and occurs at the expected mass ratio, which approximately follows the mean BH mass ($\sim 6.5 \unit{M_\odot}$) and NS mass ($\sim 1.2 \unit{M_\odot}$). Yet we also see a second peak at higher mass ratios around $q = 0.34$, which arise from the fraction of the population that underwent only stable mass transfer phases. The stability of mass transfer is strongly dependent on the mass ratio and thus these systems have a bias for more equal mass ratio systems, leading to a peak at higher $q$.

@@

The peak at recent times is due to the fact that most binaries in our sample are formed with merger times on the order of a couple of Gyr (since most form through a common-envelope phase that significantly tightens the binary). For a binary to be in the LISA band and detectable it must be near the end of its inspiral and so LISA will be most sensitive to sources that were formed a couple of Gyr ago.

@@

In the lower panel of Fig.~\ref{fig:detectable_distance_dist}, we show the density distribution for detectable DCOs in the galaxy. We see that most sources are concentrated in the Galactic center, with a bias towards the position of the Sun. Yet we also show that systems are detectable out to the edges of the galaxy. This is especially true for NSNSs, as we highlight by plotting the 75\% contour for the individual DCO types.

### 2021-07-30 paper/tex/sections/variations.tex
add hyphen to CE

In models \modBetaLow{}-\modBetaHigh{}, we set the mass transfer efficiency $\beta$ to a fixed value. For the BHBHs and BHNSs, as we increase $\beta$ the detection rates steadily decrease. This may seem unintuitive since one may assume that a higher mass transfer efficiency would lead to more massive compact objects and thus a more detectable population. However, this contributes to the envelope mass and so without increasing the core mass or fallback rate, the final compact object mass is relatively unaffected. Moreover, one must consider that most of these DCOs are formed through a common envelope event and so retaining more of the envelope during mass transfer means that the eventual ejection of the envelope is much more difficult, thus leading to more stellar mergers and fewer detectable systems \citep[e.g.][]{Kruckow+2018}.

In contrast, for the NSNSs, the detection rate increases with increasing $\beta$. This is for two main reasons: firstly the ejection of a common envelope is less problematic for NSNSs as they are less massive \citep[e.g.][]{Kruckow+2018}. Moreover, the increased mass transfer efficiency means that systems that were previously below the mass necessary to become a NS can now accrete enough mass to form a NS. Although the same is true for more massive stars becoming BHs instead of NSs, due to the IMF, there is a net flux of more stars becoming NSs.

Enforcing that case BB mass transfer is always unstable (model \modCaseBB{}) has little effect on the BHBH detection rate whilst moderately and drastically decreasing the detection rates of BHNSs and NSNSs respectively. This is because a large fraction of NSs (and the majority of NSNS binaries are formed) through case BB mass transfer. Therefore, setting this mass transfer to be always unstable results in many of these binaries to merge before they could become DCOs since we assume the pessimistic common envelope (CE) scenario by default.

@@

\subsubsection{Common envelope variation trends}\label{sec:detection_rate_CE_trends}

>>>>

In models \modBetaLow{}-\modBetaHigh{}, we set the mass transfer efficiency $\beta$ to a fixed value. For the BHBHs and BHNSs, as we increase $\beta$ the detection rates steadily decrease. This may seem unintuitive since one may assume that a higher mass transfer efficiency would lead to more massive compact objects and thus a more detectable population. However, this contributes to the envelope mass and so without increasing the core mass or fallback rate, the final compact object mass is relatively unaffected. Moreover, one must consider that most of these DCOs are formed through a common-envelope event and so retaining more of the envelope during mass transfer means that the eventual ejection of the envelope is much more difficult, thus leading to more stellar mergers and fewer detectable systems \citep[e.g.][]{Kruckow+2018}.

In contrast, for the NSNSs, the detection rate increases with increasing $\beta$. This is for two main reasons: firstly the ejection of a common-envelope is less problematic for NSNSs as they are less massive \citep[e.g.][]{Kruckow+2018}. Moreover, the increased mass transfer efficiency means that systems that were previously below the mass necessary to become a NS can now accrete enough mass to form a NS. Although the same is true for more massive stars becoming BHs instead of NSs, due to the IMF, there is a net flux of more stars becoming NSs.

Enforcing that case BB mass transfer is always unstable (model \modCaseBB{}) has little effect on the BHBH detection rate whilst moderately and drastically decreasing the detection rates of BHNSs and NSNSs respectively. This is because a large fraction of NSs (and the majority of NSNS binaries are formed) through case BB mass transfer. Therefore, setting this mass transfer to be always unstable results in many of these binaries to merge before they could become DCOs since we assume the pessimistic common-envelope (CE) scenario by default.

@@

\subsubsection{Common-envelope variation trends}\label{sec:detection_rate_CE_trends}
